---
title: Quarto Cheatsheet                      # Title of the document

# Bibliography Details
bibliography: QuartoCheatsheetReferences.bib  # Name of bib file in the same folder 
bibliographystyle: apa                        # Bibliography style

execute:                                      # Don't print warnings or messages 
  warning: false                              # in the file 
  message: false

format:
  html:                                       # html details
    number-sections: TRUE
    code-fold: TRUE
    code-line-numbers: TRUE
    cap-location: top 
  pdf:                                        # pdf details
    documentclass: article
    number-sections: TRUE
    reference-section-title: "References" 
    code-line-numbers: TRUE
    cap-location: top 
    include-in-header:
      - text: |
          \usepackage{amsmath}
          \usepackage[dvipsnames]{xcolor}
          \usepackage[shortlabels]{enumitem}
          \usepackage{parskip}
          \usepackage{hyperref}
          \usepackage{geometry}
          \usepackage{float}
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
format-links: [pdf]
---

<!-- ############### Code to adjust code block size ############### -->
```{r}
#| echo: false

default_chunk_hook <- knitr::knit_hooks$get("chunk")

latex_font_size <- c(
  "Huge",
  "huge",
  "LARGE",
  "Large",
  "large",
  "normalsize",
  "small",
  "footnotesize",
  "scriptsize",
  "tiny"
)

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if (options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
  } else {
    x
  }
})
```

<!-- ############### Start Document ############### -->
# Introduction <!-- First Section -->

Throughout the lab, we should provide an overarching narrative of what the problem is and what we are doing to solve it. Each section and subsection should be clearly and concisely communicated. You're telling a story. Each lab should be a writing sample you would be happy to submit to a hiring manager, demonstrating that you can handle complex tasks *and* communicate them in simple, concise, and easy-to-understand documents.

By default, this document will compile to an HTML file when you click the "Preview" button. To render the entire project and make it so the PDF button works, run the following code in the console. Note, you must make sure your console is running in the correct working directory. 
```{r}
#| eval: false
#| size: scriptsize
# Use the following code in the console, changing the filename to be your actual filename:
quarto::quarto_render("QuartoCheatsheet.qmd")
```
If it doesn't work, you may not be in the correct directory. Press Ctrl + Shift + p or Command + Shift + p, search the command menu for "Set as Working Directory in Active Console", navigate to the folder containing your project and click "Set Directory."

## Citations <!-- First Subsection -->

Here, I have a hidden `R` chunk for loading libraries by setting `echo: false` in the header. This chunk is not printed, so we need to cite the packages we use. For example, we use `R` [@R] for all analyses, including the `tidyverse` [@tidyverse] package for data wrangling and visualization, and data from the `palmerpenguins` package [@palmerpenguins]. Your work should include citations for *all* `R` packages you use and for `R` itself.
<!-- Code to load libraries -->
```{r}
#| label: load-packages
#| echo: false
library(tidyverse)
library(palmerpenguins)
```

## Subsection on Math Mode
How does mathematics work? We can do math inline using dollar signs (e.g., $x^2$ is a cool function). We can do a centered one-line equation using double dollar signs. 
$$f(x) = x^2 + 2$$ 
We can align an equation using the `align` \LaTeX environment. Note that we can nicely justify our work using the tag environment. 
\begin{align*}
f(x) = x^2 + 2 &\stackrel{set}{=} 0\\
x^2 &= -2 \tag{ut oh}
\end{align*}

# Data and Results

Present your data, the results of your analyses, and how each reported aspect contributes to answering the question. This section should include table(s), statistic(s), and graphical displays.

Tie the data and the results to the introduction, where you introduce the problem at hand and what you propose to do to answer the question. Make sure to present the results in a sensible order and that each result contributes to a logical, well-developed solution. I like using subsections to concisely make an argument for each substantive point I need to address. Avoid being repetitive.

## Figures

We can create and load graphs like @fig-box-billlength into our documents as follows. Note that when inserting a figure, your must label the `R` code chunk using the prefix `fig-`.

```{r}
#| label: fig-box-billlength
#| fig-cap: "Bill length (mm) by species."
#| fig-width: 5
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: scriptsize

ggplot(data = penguins) +
  geom_boxplot(aes(y = bill_length_mm, x = species)) +
  theme_bw() +
  labs(x = "Species", y = "Bill Length (mm)")
```

## Tables

We can create and load graphs like @tbl-nsum-billlength into our documents as follows. Note that when inserting a table, your must label the `R` code chunk using the prefix `tbl-`.

```{r}
#| size: scriptsize
penguins.summary <- penguins |>
  group_by(species) |>
  summarize(
    n = n(),
    Missing = sum(is.na(bill_length_mm)),
    Mean = mean(bill_length_mm, na.rm = T),
    SD = sd(bill_length_mm, na.rm = T),
    Median = median(bill_length_mm, na.rm = T),
    IQR = IQR(bill_length_mm, na.rm = T)
  ) |>
  rename(
    "Species" = "species",
  )
```

We can print the table nicely using the `kable()` function. Note that the code to print the table nicely is hidden by setting `echo: false` in the header.
```{r}
#| label: tbl-nsum-billlength
#| tbl-cap: "Bill length (mm) by species."
#| tbl-align: "center"
#| echo: false
penguins.summary |>
  knitr::kable(digits = 2)
```

# Discussion

This is where we tie all the work together into a concise, clear, and camera-ready answer. What is the takeaway?

This does not mean you should write something like this.

> *I did this, and then I did that, and then I did this other thing, and then..., and then...,and then...*

Instead, provide a clear and concise narrative that flows from the problem specification in the Introduction to how you have answered it.

You should objectively evaluate the evidence you found in the data. Do not embellish or wish-terpet (my made-up phrase for making an interpretation you, or the researcher, wants to be true without the data actually supporting it). Connect your findings to the existing information you provided in the Introduction.

Finally, provide some concluding remarks that tie together the entire paper. Think of the last part of the results as abstract-like. Tell the reader what they just consumed -- what's the takeaway message?

